Hypertension and microbiome

### 2019-08-26 articletwo.Rmd

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(pwr)
library(gtable)
library(car)
library(M3C)
```

@@

Loading phyloseq object

@@

 ```{r adonis calculation}
 if (!file.exists("rds/adonis.species.rds")) {
     adonis.species = list()
     adonis.species$min <- calculateadonis(dset = meta(pseq.species),
                                           matrix = bray.dist.m.species,
                                           covariates = var.CL.min,
                                           npermutations = 9999)
     adonis.species$max <- calculateadonis(dset = meta(pseq.species),
                                           matrix = bray.dist.m.species,
                                           covariates = var.CL,
                                           npermutations = 9999)
     saveRDS(adonis.species, file = "rds/adonis.species.rds")
 } else {
     adonis.species  <- readRDS("rds/adonis.species.rds")
 }
```

@@

```{r alphabeta definitions}
dset.alpha <- meta.merge.alphadiversity(pseq.species) %>%
    mutate(diversity_shannon = scale(diversity_shannon),
           MAP = oMAP,
           PULSEPRESSURE = oPULSEPRESSURE,
           SYSTM = oSYSTM,
           DIASM = oDIASM)

alphadiversity <- list(
    min = calculateglm(dset.alpha,
                       modelstring = sprintf("%%s ~ %s + diversity_shannon",
                                             paste0(var.CL.min, collapse = " + ")),
                       filterstr = "shannon"),
    max = calculateglm(dset.alpha,
                       modelstring = sprintf("%%s ~ %s + diversity_shannon",
                                             paste0(var.CL, collapse = " + ")),
                       filterstr = "shannon"))

@@

```{r save grob}
alpha.diversity.min <- plot.alpha(diversities$min, alphabreaks = seq(-1,0,0.2))
alpha.diversity.max <- plot.alpha(diversities$max, alphabreaks = seq(-1,0,0.2))

beta.diversity.min <- plot.beta(diversities$min)
beta.diversity.max <- plot.beta(diversities$max)

gs4 <- list(alpha.diversity.min$yaxis,
            alpha.diversity.min$plot,
            beta.diversity.min$plot,
            alpha.diversity.min$legend,
            alpha.diversity.min$xaxis,
            beta.diversity.min$xaxis,
            beta.diversity.min$xlab,
            text_grob("A.", size = 18),
            alpha.diversity.max$yaxis,
            alpha.diversity.max$plot,
            beta.diversity.max$plot,  
            alpha.diversity.max$xaxis,
            beta.diversity.max$xaxis,
            beta.diversity.max$xlab,
            text_grob("B.", size = 18))

g4 <- arrangeGrob(grobs = gs4,
                   layout_matrix = rbind(
                      c(8,    NA,    NA, NA,   NA, NA, NA),
                      c(NA, 1,    2,    NA, 3, NA, NA),
                      c(NA, 1,    2,    NA, 3, NA, 4),
                      c(NA, NA, 5,    NA,6, NA, NA),
                      c(NA, NA, NA, NA,7, NA, NA),
                      c(15,    NA,    NA, NA,   NA, NA, NA),
                      c(NA, 9,    10,  NA,  11, NA, NA),
                      c(NA, 9,    10,  NA,  11, NA, NA),
                      c(NA, NA, 12,   NA, 13, NA, NA),
                      c(NA, NA, NA, NA,14, NA, NA)),
                  widths=c(0.7, 4, 0.7, 0.1, 5, 0.7, 3),
                  heights=rep(c(1, 1, 6, 1, 2), 2))

ggsave(file = "cache/alpha-beta.png", plot=g4, height=6, width=9)
```

@@

plot_labeller <-  function(value) {
        plot.labels <- list("Axis.2" = sprintf("A."),
                            "Axis.3" = sprintf("B."))
        return(plot.labels[value])
}

pcoa.plot <- ggplot(data=pcoa.df %>% gather(key, Axis, Axis.2, Axis.3),
                    aes(x=Axis.1, y=Axis, color = HYPERTENSION)) +
    facet_wrap(~key, scales = "free_y", strip.position = "left", labeller = axis_labeller) +
    geom_point(size=0.1, alpha=0.4) +
    xlab(sprintf("PCoA Axis 1 (%.1f%%)", 100*eig.fracs[1])) +
    ylab(NULL) +
    scale_colour_manual(name = "State and treatment",
                        labels = c("Normotensive", "Hypertensive"),
                        breaks=c(0, 1),
                        values = c("blue", "red")) +   
    scale_x_continuous(breaks=seq(-5, 5, 1)) +
    scale_y_continuous(breaks=seq(-5, 5, 1)) +
    theme_classic() +
    guides(colour = guide_legend(override.aes = list(size=4))) +
    coord_cartesian(clip = 'off') +
    geom_text(aes(x = -Inf, y = Inf, label = plot_labeller(key)),
              size = 5,
              hjust = 2,
              vjust = -1,
              inherit.aes = FALSE) +
    theme(text = element_text(size=12),
          legend.position = "none",
          strip.background = element_blank(),
          strip.placement = "outside",
          aspect.ratio = 1,
          strip.text = element_text(size = 14),
          axis.title.x = element_text(size = 14))

@@

```{r top generas, eval = FALSE, include = FALSE}
abund.genus <- abundances(pseq.genus, transform = "compositional") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id") %>%
    setNames(tolower(replace.brackets(names(.))))

abund.genus.bacteria <- abund.genus %>% select(-contains("virus")) %>% colnames

abund.df <- merge(pcoa.vectors %>%
                  select("Axis.1", "Axis.2", "Axis.3") %>%
                  tibble::rownames_to_column("id"),
                  abund.genus %>% select(id, abund.genus.bacteria),
                  by = "id") %>%
    merge(.,  meta(pseq.genus) %>% tibble::rownames_to_column("id"), by = "id")

pcoacorabund <- function(df,
                         genera.names,
                         axis.names = c("Axis.1", "Axis.2", "Axis.3")) {
    lapply(c2l(genera.names), function (taxa) 
        lapply(c2l(axis.names), function(axis) {
            fo <- sprintf("%s ~ %s", axis, taxa)
            model <- lm(as.formula(fo), data = df)
            summary(model)$r.squared
        }))
}

pcoacor.genus <- pcoacorabund(abund.df, genera.names = abund.genus.bacteria)
pcoacor.genus.df <- map_df(pcoacor.genus, ~as.data.frame(.x), .id="id")

pcoacor.genus.df %>%
    arrange(-Axis.3) %>%
    head(20) %>% kable
```

@@

```{r pcoa analysis main axis}
glm.pcoa <- lapply(c2l(paste0("Axis.", seq(1,5))), function(axis)
    calculateglm(pcoa.df,
                 modelstring = sprintf("%%s ~ %s + %s", axis, paste(var.CL, collapse = "+")),
                 responses = list(model_1 = "oMAP", model_2 = "oSYSTM", model_3 = "oDIASM",
                                          model_4 = "oPULSEPRESSURE", model_5 = "HYPERTENSION"),
                 filterstr = "Axis"))

do.call(rbind, glm.pcoa) %>% 
    select(response, term, estimate, std.error, p.value, conf.low, conf.high) %>%
    mutate_if(is.numeric, round, 3) %>%
    arrange(p.value) %>%
    kable

@@

```{r lacto adjusted means}
df.genus.salt.unadjusted <- with(df.genus.salt, data.frame(salt_groups = as.factor(seq(3)),
                                                           SEX = as.factor(1),
                                                           BL_AGE = mean(BL_AGE)))

adjmean.lacto.saltgroup <- predict(lm.lacto.saltgroup,
                               df.genus.salt.unadjusted,
                               interval = "prediction") %>%
    as.data.frame %>%
    dplyr::rename(estimate = fit,
                  conf.low = lwr,
                  conf.high = upr) %>%
    tibble::rownames_to_column("term")
```

@@

```{r forest plot lacto}
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Dietary salt",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(3.0, 8.0), breaks = seq(0, 10, 1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.4, 1.5), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)

ggsave(file = "cache/salt-forest.png", plot=gsalt, height=3, width=5, dpi = 300)
```

@@

@@

>>>>

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(pwr)
library(gtable)
library(car)
library(M3C)
library(emmeans)
```

@@

eLoading phyloseq object

@@

 ```{r adonis calculation}
 if (!file.exists("rds/adonis.species.rds")) {
     adonis.species <- calculate.betadiversity(pseq = pseq.species,
                                               matrix = bray.dist.m.species,
                                               vars = list("max" = var.CL,
                                                            "min" = var.CL.min))

     saveRDS(adonis.species, file = "rds/adonis.species.rds")
 } else {
     adonis.species  <- readRDS("rds/adonis.species.rds")
 }
```

@@

```{r alphabeta definitions}
alphadiversity <- calculate.alphadiversity(pseq.species,
                                           list("max" = var.CL,
                                                "min" = var.CL.min))

@@

```{r save grob}
g4 <- plot.diversities(diversities)
ggsave(file = "cache/alpha-beta.png", plot=g4, height=6, width=9)
```

@@

pcoa.plot <- ggplot(data=pcoa.df %>% gather(key, Axis, Axis.2, Axis.3),
                    aes(x=Axis.1, y=Axis, color = HYPERTENSION)) +
    facet_wrap(~key, scales = "free_y", strip.position = "left", labeller = axis_labeller) +
    geom_point(size=0.1, alpha=0.4) +
    xlab(sprintf("PCoA Axis 1 (%.1f%%)", 100*eig.fracs[1])) +
    ylab(NULL) +
    scale_colour_manual(name = "State and treatment",
                        labels = c("Normotensive", "Hypertensive"),
                        breaks=c(0, 1),
                        values = c("blue", "red")) +   
    scale_x_continuous(breaks=seq(-5, 5, 1)) +
    scale_y_continuous(breaks=seq(-5, 5, 1)) +
    theme_classic() +
    guides(colour = guide_legend(override.aes = list(size=4))) +
    coord_cartesian(clip = 'off') +
    theme(text = element_text(size=12),
          legend.position = c(0.35, 0.95),
          legend.title = element_blank(),
          legend.text = element_text(size = 10),
          legend.key.size = unit(5, "mm"),
          legend.background = element_blank(),
          strip.background = element_blank(),
          strip.placement = "outside",
          aspect.ratio = 1,
          strip.text = element_text(size = 14),
          axis.title.x = element_text(size = 14))

@@

```{r top pcoa genera}
df.genus.abund <- abundances(pseq.genus, transform = "compositional") %>%
    as.data.frame %>%
    tibble::rownames_to_column() %>%
    dplyr::mutate(rowname = make.names(rowname)) %>%
    gather(id, value, -rowname) %>% 
    spread(rowname, value)

df.genus.pcoa <- merge(pcoa.vectors %>% tibble::rownames_to_column("id"),
                         df.genus.abund,
                         by = "id")

pcoacorabund <- function(df,
                         taxas,
                         axis.names = c("Axis.1", "Axis.2", "Axis.3")) {
    mclapply(c2l(taxas), function (taxa) 
        lapply(c2l(axis.names), function(axis) {
            fo <- sprintf("%s ~ %s", axis, taxa)
            model <- lm(as.formula(fo), data = df)
            summary(model)$r.squared
        }), mc.cores = 20)
}

pcoa.correlations <- pcoacorabund(df.genus.pcoa, taxas = colnames(df.genus.abund)[-1])

pcoa.correlations.df <- map_df(pcoa.correlations, ~as.data.frame(.x), .id="id")

pcoa.correlations.df %>% head

pcoa.top.names <- lapply(c2l("Axis.1", "Axis.2", "Axis.3"), function(x) {
    top.names <- pcoa.correlations.df %>% arrange(desc(!!sym(x))) %>% head(10) %>% pull(id)
    pcoa.correlations.df %>%
        dplyr::filter(id %in% top.names) %>%
        select(id, x) %>%
        tibble::rownames_to_column(var = "number") %>%
        dplyr::mutate(!! sprintf("taxa_%s", x) := gsub("g_", "", id)) %>%
        select(one_of(sprintf("taxa_%s", x), x, "number"))
    })

pcoa.top.table <- pcoa.top.names %>%
    reduce(left_join, by = "number") %>%
    select(-number) %>%
    mutate_if(is.numeric, round, 3)
```

@@

```{r pcoa analysis main axis}
glm.pcoa <- lapply(c2l(paste0("Axis.", seq(1,3))), function(axis)
    calculateglm(pcoa.df,
                 modelstring = sprintf("%%s ~ %s + %s", axis, paste(var.CL, collapse = "+")),
                 responses = list(model_1 = "oMAP", model_2 = "oSYSTM", model_3 = "oDIASM",
                                          model_4 = "oPULSEPRESSURE", model_5 = "HYPERTENSION"),
                 filterstr = "Axis"))

do.call(rbind, glm.pcoa) %>% 
    select(response, term, estimate, std.error, p.value, conf.low, conf.high) %>%
    dplyr::mutate(Qval = p.adjust(p.value, method="BH")) %>%
    mutate_if(is.numeric, round, 3) %>%
    arrange(Qval) %>%
    kable

@@

```{r lacto adjusted means}
adjmean.lacto.saltgroup <- emmeans::emmeans(lm.lacto.saltgroup, specs = c("salt_groups")) %>%
    as.data.frame %>%
    dplyr::rename(term = salt_groups, estimate = emmean, conf.low = lower.CL, conf.high = upper.CL)
```

@@

```{r forest plot lacto}
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Tertile of 24-hour urinary sodium",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(5.4, 5.65), breaks = seq(0, 10, 0.1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Tertile of Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.4, 1.5), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)

ggsave(file = "cache/salt-forest.png", plot=gsalt, height=3, width=5.5, dpi = 300)
```

@@

Testing salt model using maaslin

```{r salt model maaslin}
taxa(pseq.genus.core) %>% mygrep(word = "lacto")

pseq.genus.salt <- prune_taxa(c("g_Lactobacillus"), pseq.genus.core)

pseq.genus.salt.meta <- meta(pseq.genus.salt) %>%
    tibble::rownames_to_column(var = "rowname") %>%
    dplyr::filter(!is.na(NA.)) %>%
    dplyr::mutate(dUNA = myscale(NA.),
                  salt_groups = relevel(factor(dplyr::ntile(dUNA, 3)), ref='1')) %>%
    tibble::remove_rownames() %>%
    tibble::column_to_rownames(var = "rowname") %>%
    select(var.BP, var.CL.min, starts_with("salt_groups"))

phyloseq::sample_data(pseq.genus.salt) <- phyloseq::sample_data(pseq.genus.salt.meta)

maaslin.salt <- maaslinwrapper(pseq = pseq.genus.salt,
                               looped = var.BP,
                               forced = c(var.CL.min, "salt_groups"),
                               taxa = "g_Lactobacillus")

maaslin.salt %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("[gs]_", "", Feature),
                  beta = Coefficient) %>%
    arrange(Feature)
```

@@

```{r write supplement}
core.names <- pseq.genus.core %>%
    taxa %>%
    gsub("g_", "", .) %>%
    as.data.frame %>%
    setNames("name") %>%
    arrange(name) %>%
    mutate(name = as.character(name)) %>%
    mutate(class = case_when(row_number()-1 < n()/4 ~ 1,
                             row_number()-1 < 2*n()/4 ~ 2,
                             row_number()-1 < 3*n()/4 ~ 3,
                             TRUE ~ 4))

core.names %>% pull(class) %>% as.factor %>% summary

core.names.fourcols <- cbind(
    core.names %>% filter(class == 1) %>% pull(name),
    core.names %>% filter(class == 2) %>% pull(name),
    core.names %>% filter(class == 3) %>% pull(name),
    core.names %>% filter(class == 4) %>% pull(name)) %>%
    as.data.frame

core.names.fourcols %>% unlist %>% length

core.names.flextable <- flextable(core.names.fourcols) %>%
    flextable::fontsize(size = 10, part = "header") %>%
    flextable::fontsize(size = 10, part = "body") %>%
    flextable::width(j=1:4, width = 1.8) %>%
    flextable::align(align = "left", part = "header") %>%
    flextable::align(align = "left")

doc.s <- read_docx(path = "style/articlestyle.docx") %>%
  body_remove()
writetable(doc.s, core.names.flextable, number = "S1", "Core genera", "") %>%
print(doc.s, target = paste0("report/articletwo-supplement.docx"))
```
